Concurrent Programming in Harmony

### 2021-12-24 book/paper.tex
Latest

\begin{itemize}
\item Variables may have more bits than the processor's data bus.
For example, variables may have 32 bits, but the data bus may only
have 16 bits.  Thus to store or load a variable takes two 16-bit
operations each.  Take, for example, a variable that has value
0xFFFFFFFF, and consider a concurrent load and store operation on
the variable.
The store operation wants to clear the variable, but because it
takes two store operations on the bus, the load operation may
return either 0xFFFF0000 or 0x0000FFFF, a value that the variable
never was supposed to have.
This is the case even if the processor supports a 32-bit load or
store machine instruction: on the data bus it is still two
operations.
\item Modern processors sometimes re-orders load and store operations
(out-of-order execution) for improved performance.  On a sequential processor,
the re-ordering is not a problem as the processor only re-orders
operations on independent memory locations.  However, as \autoref{ex:reorder}
showed, Peterson's algorithm breaks down if operations are re-ordered.
Some memory caches can also cause non-atomic behavior of memory when shared
among multiple cores.
\item Even compilers, in their code generation, may make optimizations
that can reorder operations, or even eliminate operations, on variables.
For example, a compiler may decide that it is unnecessary to read the same
variable more than once, because how could it possibly change if there
are no store operations?
\end{itemize}

Peterson's algorithm relies on a \emph{sequential consistent memory model}
and hence the \textbf{sequential} statement: without it Harmony will
complain about data races.
More precisely, the \textbf{sequential} statement says that the
program relies on memory load and store instructions operating on
the indicated variables to be performed sequentially, and that this
order should be consistent with the order of operations invoked on each
thread.
The default memory models of C and Java are not sequentially
consistent.
The unfortunately named \texttt{volatile} keyword in Java has
a similar function as Harmony's \texttt{sequential} keyword.
Like many constructions in Java, its \texttt{volatile} keyword
was borrowed from C and C++.
However, in C and C++ they do \emph{not} provide sequential
consistency, and one cannot implement Peterson's algorithm in
C or C++ directly.

>>>>

\begin{itemize}
\item Variables may have more bits than the processor's data bus.
For example, variables may have 32 bits, but the data bus may only
have 16 bits.  Thus to store or load a variable takes two 16-bit
operations each.  Take, for example, a variable that has value
0xFFFFFFFF, and consider a concurrent load and store operation on
the variable.
The store operation wants to clear the variable, but because it
takes two store operations on the bus, the load operation may
return either 0xFFFF0000 or 0x0000FFFF, a value that the variable
never was supposed to have.
This is the case even if the processor supports a 32-bit load or
store machine instruction: on the data bus it is still two
operations.
\item Modern processors sometimes re-orders load and store operations
(out-of-order execution) for improved performance.  On a sequential processor,
the re-ordering is not a problem as the processor only re-orders
operations on independent memory locations.  However, as \autoref{ex:reorder}
showed, Peterson's algorithm breaks down if such seemingly independent
operations are re-ordered.
Some memory caches can also cause non-atomic behavior of memory when shared
among multiple cores.
\item Even compilers, in their code generation, may make optimizations
that can reorder operations, or even eliminate operations, on variables.
For example, a compiler may decide that it is unnecessary to read the same
variable more than once, because how could it possibly change if there
are no store operations in between?
\end{itemize}

Peterson's algorithm relies on a \emph{sequential consistent memory model}
and hence the \textbf{sequential} statement: without it Harmony will
complain about data races.
More precisely, the \textbf{sequential} statement says that the
program relies on memory load and store instructions operating on
the indicated variables to be performed sequentially, and that this
order should be consistent with the order of operations invoked on each
thread.
The default memory models of C and Java are not sequentially
consistent.
The unfortunately named \texttt{volatile} keyword in Java has
a similar function as Harmony's \texttt{sequential} keyword.
Like many constructions in Java, its \texttt{volatile} keyword
was borrowed from C and C++.
However, in C and C++, they do \emph{not} provide sequential
consistency, and one cannot implement Peterson's algorithm in
C or C++ directly.
